Uncertainty in social learning evolution

### 2022-05-22 ms.tex
updated schematic and wrote first analysis paragraph

Social learning is essential to human and other species' everyday life and survival.
It allows individuals to solve problems when acquiring information from others is
more efficient than learning on one's own~\cite{Laland2004}. Theory predicts that
social learning should be favored in contexts with greater
uncertainty~\cite{BoydRicherson1985,Henrich1998}, up to a
point~\cite{Rogers1988,Feldman1996}; this prediction has received some empirical
support across species~\cite{McElreath2005,Kendal2018,Allen2019}.  However, the
meaning of the term \emph{uncertainty} often conflates environmental variability,
spatial heterogeneity, ambiguity or uncertainty about payoff structure, and other
variability that could be plausibly interpreted as \emph{uncertainty}.  Many models
of social learning evolution do not account for 
individual-level cognition~\cite{Heyes2016}, though humans
clearly have evolved cognitive mechanisms for dealing with 
uncertainty generally~\cite{Gershman2019,Schulz2019}.
Furthermore, social learning evolution theory tends to be organized around an
increasing number of modeling choices~\cite[Figure 1]{Kendal2018} instead
of around broader contextual assumptions such as uncertainty conditions. It remains
an outstanding question, then, to identify how different contexts and forms of
uncertainty might combine to affect social learning evolution in complex ways.
Answering this question is critical for understanding human behavior adapting to a
rapidly changing world, both to mitigate existential
threats~\cite{Moya2020,Jones2021}, and to captialize on new opportunities such as
transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.
We will answer this question about how uncertainty drives evolution, including
identifying regimes where evolution itself is highly uncertain and often 
leads populations into suboptimal learning strategies~\cite{Rogers1988}.

@@

To answer our research questions
we developed an agent-based model of social learning evolution in which a society of
$N$ individuals each must decide which of $B$ behaviors to perform at each time step
within a generation of $L$ time steps. At the end of each generation, agents are selected
with replacement from the population to reproduce another $N$ agents. Agents who
inherit the social learning trait then learn about behavioral payoffs from the
previous generation, while asocial learners begin life with no \emph{a priori}
knowledge of the world.  Each behavior is a ``bandit'', a common modeling and
experimental approach for representing behaviors with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Rendell2010,Schulz2019}.  Each behavior
$b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; $\pi_b$ is therefore also the expected payoff of
behavior $b$. One behavior in each generation is optimal and pays off $\pihigh$,
while the rest pay off $\pilow < \pihigh$.  Model agents must decide which behavior to
perform at each time step.  To do this, agents use an explore-exploit strategy to
sometimes try the most profitable behavior they know about, and other times try
alternatives that may pay off more reliably.  We developed a series of computational
analyses by systematically varying uncertainty parameters and observing the
frequency that model populations evolve to be social learners, the average payoffs
accrued by agents in each setting, and the time it takes for model populations to
reach fixation (agents all social or all asocial learners).

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied their values and observed three key outcome variables aggregated across 1000
trial simulations, and across all agents. We varied $u \in \{0.0,
0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L
\in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  We observed
three outcome measures: (1) $\meansl$, the average value of $s_i$ over all agents
and trials; (2) $\meanpi / L$, the average net payoffs within agent lifetimes over
trials, normalized by their lifespan; and (3) $\meanT / L$, or the mean number
of time steps to fixation normalized by lifespan\footnote{Equivalent to the mean
number of generations to fixation} 
(fixation means $\sum_i s_i = 0 \text{ or } 1$). \mt{Make sure to qualitatively
introduce these and their evolutionary importance in the intro.}
We inspect the outcomes through visualization environmental
variation is on the x-axis since it is the best understood: we expect that as $u$
increases, social learning prevalence $\meansl \to 0$. We expected and observed
that the other three uncertainty factors shift at which $u$ $\meansl$ starts to
decrease, and how rapidly $\meansl$ decreases as $u$ increases. Note the
hypothesis-testing concept of significance is meaningless here because we could
make any small outcome difference significant by running more simulation trials.
The important outcomes for this study are the patterns of our outcome 
variables over the systematic variation of the uncertainty factors.

@@

@@

\paragraph{Analysis 2: evolutionary uncertainty when $\meansoc \approx \meanasoc$}

\begin{figure}
  \caption{Mean payoff normalized for lifespan (y-axes, solid line with circles)
    generally track either expected all-social-learners payoff ($\meansoc$, dash-dot
    line with diamonds) or expected all-asocial payoff ($\meanasoc$, long dash
    horizontal lines) as $u$ increases (x-axes). When $\meansl$ goes from 1 to
  0 is when social learning payoffs approach individual learning payoffs due to
increased environmental variability.} 
  \label{fig:netPayoffs}
\centering
    \includegraphics[width=0.75\textwidth]{Figures/meanNetPayoffs.pdf}
\end{figure}

>>>>

Social learning is essential to human and other species' everyday life and survival.
It allows individuals to solve problems when acquiring information from others is
more efficient than learning on one's own~\cite{Laland2004}. Theory predicts that
social learning should be favored in contexts with greater
uncertainty~\cite{BoydRicherson1985,Henrich1998}, up to a
point~\cite{Rogers1988,Feldman1996}; this prediction has received some empirical
support across species~\cite{McElreath2005,Kendal2018,Allen2019}.  However, the
meaning of the term \emph{uncertainty} often conflates environmental variability,
spatial heterogeneity, ambiguity or uncertainty about payoff structure, and other
variability that could be plausibly interpreted as \emph{uncertainty}.  Many models
of social learning evolution do not account for 
individual-level cognition~\cite{Heyes2016}, though humans
clearly have evolved cognitive mechanisms for dealing with 
uncertainty generally~\cite{Gershman2019,Schulz2019}.
Furthermore, social learning evolution theory tends to be organized around an
increasing number of modeling choices~\cite[Figure 1]{Kendal2018} instead
of around broader contextual assumptions such as uncertainty conditions. It remains
an outstanding question, then, to identify how different contexts and forms of
uncertainty might combine to affect social learning evolution in complex ways.
Answering this question is critical for understanding human behavior adapting to a
rapidly changing world, both to mitigate existential
threats~\cite{Moya2020,Jones2021}, and to captialize on new opportunities such as
transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.
We will answer this question about how uncertainty drives social learning 
evolution, including
identifying regimes where evolution itself is highly uncertain and often 
leads populations into suboptimal learning strategies~\cite{Rogers1988}.

@@

To answer our research questions we developed an agent-based model of social
learning evolution in which a society of $N$ individuals each must decide which of
$B$ behaviors to perform at each time step within a generation of $L$ time steps. At
the end of each generation, agents are selected with replacement from the population
to reproduce another $N$ agents. Agents who inherit the social learning trait then
learn about behavioral payoffs from the previous generation, while asocial learners
begin life with no \emph{a priori} knowledge of the world.  Each behavior is a
``bandit'', a common modeling and experimental approach for representing behaviors
with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Rendell2010,Schulz2019}.  Each behavior
$b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; $\pi_b$ is therefore also the expected payoff of
behavior $b$. One behavior in each generation is optimal and pays off $\pihigh$,
while the rest pay off $\pilow < \pihigh$.  Model agents must decide which behavior
to perform at each time step.  To do this, agents use an explore-exploit strategy to
sometimes try the most profitable behavior they know about, and other times try
alternatives that may pay off more reliably.  We developed a series of computational
analyses by systematically varying uncertainty parameters and observing the
frequency that model populations evolve to be social learners, the average payoffs
accrued by agents in each setting, and the time it takes for model populations to
reach fixation (agents all social or all asocial learners). Model dynamics are
illustrated in Figure~\ref{fig:schematic}.

\begin{figure}
  \caption{Agents are initialized as randomly social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs (1), which goes on for $L$
timesteps (B). After $L$ time steps (2), agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). Stop the simulation if children are all
social or asocial learners, otherwise repeat another generation and evolution (3).}
  \label{fig:schematic}
  \centering
    \includegraphics[width=\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied their values and observed three key outcome variables aggregated across 1000
trial simulations, and across all agents. We varied $u \in \{0.0,
0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L
\in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  We observed
three outcome measures: (1) $\meansl$, the average value of $s_i$ over all agents
and trials; (2) $\meanpi / L$, the average net payoffs within agent lifetimes over
trials, normalized by their lifespan; and (3) $\meanT / L$, or the mean number
of time steps to fixation normalized by lifespan\footnote{Equivalent to the mean
number of generations to fixation} 
(fixation means $\sum_i s_i = 0 \text{ or } 1$). \mt{Make sure to qualitatively
introduce these and their evolutionary importance in the intro.}
We inspect the outcomes through by plotting $\meansl$ on the y-axis and environmental
variation is on x-axis, since it is the best understood: we expect that as $u$
increases, $\meansl$ starts at 1 and goes to 0. We expected and observed
that the other three uncertainty factors shift at which $u$ $\meansl$ starts to
decrease, and how rapidly $\meansl$ decreases as $u$ increases. Note that the
hypothesis-testing concept of significance is meaningless here because we could
make any small outcome difference significant by running more simulation trials.
The important outcomes for this study are the patterns of our outcome 
variables over the systematic variation of the uncertainty factors. 
Our model was implemented in the Julia programming language~\cite{Bezanson2017} 
using the Agents.jl agent-based modeling library~\cite{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. Model code and
data are publicly available on GitHub\footnote{\url{https://github.com/mt-digital/UncMod}} 
and OSF.io\footnote{\mt{TODO}}.

@@

We observed several patterns in the suppression of social learning as environmental
variability, $u$, increased across different uncertainty contexts defined by the
payoff ambiguity, selection set size, and effective lifespan
(Figure~\ref{fig:mainResults}). When payoff ambiguity and selection set size
were both small ($\pilow = 0.1$ and $B=2$) social learning only evolved when
$u$ is small since individual learning is relatively efficient in this case
(Figure~\ref{fig:mainResults} upper left). As $\pilow$ increased the suppression of
$\meansl$ flattened over $u$, which was more pronounced for smaller $L$ (rows of
Figure~\ref{fig:mainResults}). 
When $L$ is smaller, agents have fewer opportunities to learn through trial and
error, and so it is more difficult for them to discern which strategy, social
or asocial, is optimal. Increased selection set size led to both a shift in 
the critical value of $u$ when $\meansl$ begins to decrease, and flattening of
this social learning suppression. The start of $\meansl$ suppression shifts
to higher $u$ since misleading social information is less consequential due to
(1) larger $B$ being a more difficult problem for trial-and-error learning and
(2) less probability mass of a misidentified optimal behavior relative to
smaller $B$, meaning agents are more likely to explore behaviors identified
as non-optimal (columns of Figure~\ref{fig:mainResults}). 
Increased $B$ led to more pronounced flattening of social learning
suppression when $L$ was small, apparently again because of the limited data
available to evolution and more difficult problem caused more drift.

@@

\paragraph{Analysis 2: how payoffs and generations to convergence support our
explanations}

\begin{figure}
  \caption{Mean payoff normalized for lifespan (y-axes, solid lines with circles)
    generally track either expected all-social-learners payoff ($\meansoc$, dash-dot
    line with diamonds) or expected all-asocial payoff ($\meanasoc$, long dash
    horizontal lines) as $u$ increases (x-axes). When $\meansl$ goes from 1 to
  0 is when social learning payoffs approach individual learning payoffs due to
increased environmental variability.} 
  \label{fig:netPayoffs}
\centering
    \includegraphics[width=0.75\textwidth]{Figures/meanNetPayoffs.pdf}
\end{figure}
